---
title: "dada2 Single-End Protocol - Combined Runs,  VMT samples"
author: "Fatima Aysha Hussain; adapted from code by Seth Bloom"
date: '`r format(Sys.Date(), "2021-05-14")`'
output:
  html_document: default
  word_document: default
---

## References:  
* Callahan, B. J., McMurdie, P. J., Rosen, M. J., Han, A. W., Johnson, A. J. A., & Holmes, S. P. (2016). DADA2: High-resolution sample inference from Illumina amplicon data. Nature Methods, 13(7), 581–3.  
    + https://doi.org/10.1038/nmeth.3869  
* McMurdie, P. J., & Holmes, S. (2013). phyloseq: an R package for reproducible interactive analysis and graphics of microbiome census data. PloS One, 8(4), e61217.  
    + https://doi.org/10.1371/journal.pone.0061217  


```{r clear-environment}
#Clears environment to remove all variables, user-defined functions, etc.
rm(list=ls()) 
```

```{r start-tracking-processing-time}
#Define variable with start time of running script.
start_time <- proc.time()
```


```{r define-sequencing-run-date-and-version}
#Define sequencing run date in format YYYY_MM_DD
print("Sequencing run date and version:")
run.date <-  "2021_05_14_v4" #Enter character string here, for example: "2018_07_25_v1"
run.date

```


```{r initiate-environment}
# Load libraries
library(ShortRead)
packageVersion("ShortRead")
library(dada2)
packageVersion("dada2")
library("phangorn")
packageVersion("phangorn")
library("phyloseq")
packageVersion("phyloseq")
library("ips")
packageVersion("ips")
library("tidyverse")
packageVersion("tidyverse")

# Set random seed for reproducibility purposes
set.seed(100)

```
###Filter and trim reads

```{r establish-directory-structure}
#Define and print path to working directory
pathwd <- "./VMT_analysis" 
print("Working directory:")
pathwd

# Define path to subdirectory containing demultiplexed forward-read fastq files
pathF <- file.path(pathwd, "Data") 
print("Directory containing raw demultiplexed .fastq sequence files:")
pathF

#Define path to directory containing 16S databases
print("Path to 16S databases:")
pathdb <- file.path(pathwd, "DADA2_taxa_dbs")
pathdb

#Define path to directory containing mapping file
print("Path to mapping file:")
pathMap <- file.path(pathwd, "Mapping")
pathMap


#Create and define path to subdirectory in which filtered files will be stored
filtpathF <- file.path(pathF, "filtered") 
print("Directory that will contain filtered .fastq files:")
filtpathF
dir.create(filtpathF)

#Create and define path to subdirectory in which to store output figures and tables
print("Directory in which to store output figures and tables:")
pathOut <- file.path(pathwd, "Output")
pathOut
dir.create(pathOut)

#Create and define path to subdirectory in which RDS files will be stored
print("Path to directory for saving RDS files:")
pathF.RDS <- file.path(pathwd, "RDS")
pathF.RDS
dir.create(pathF.RDS)

#Create and define path to subdirectory in which RDS files containing phyloseq objects will be stored
print("Path to sub-directory in which to save phyloseq objects:")
pathps0 <- file.path(pathwd, "ps_objects")
pathps0
dir.create(pathps0)
```


```{r Define-sequence-file-names-and-sample-names}
# File parsing
#Create vector of fastq files in pathF.directory
#If analyzing paired end reads, the search pattern below would need to be modified (see DADA2 tutorial)
print("Sequence files to be analyzed:")
fnFs <- sort(list.files(pathF, pattern="fastq", full.names = TRUE))
head(fnFs)

#Extract sample names, assuming filenames have format: SAMPLENAME.fastq or SAMPLENAME.fastq.gz
sample.names <- sapply(strsplit(basename(fnFs), ".fastq"), `[`, 1)
head(sample.names)
```


```{r plot-unfiltered-run-quality}
#Plot forward read quality aggregated across all forward read files
#If analyzing paired end data, would need additional code for plotting reverse read quality
p.qual.f.1.agg <- plotQualityProfile(fnFs, aggregate = TRUE) + 
  ggtitle("Fwd read quality aggregated across samples")
p.qual.f.1.agg

#Save figure to file
ggsave(filename = file.path(pathOut, paste0("Read_quality_aggregate_fwd_", run.date, ".pdf")), 
       plot = p.qual.f.1.agg, device = "pdf", width = 8, height = 6, units = "in")

```


```{r calculate-quality-plot-processing-time}
#Calculate cumulative processing time
print("Cumulative processing time (seconds):")
proc.time() - start_time
```

##Define positions at which to trim filtered data


```{r define-trim-positions}
FwdTrimLeft <-   10 #Need to insert a value here. Our typical default values is 10
FwdTrimRight <-  230 #Need to insert a value here. Typical default value is 230.

print(paste0("Trimming forward reads on left at base position ", FwdTrimLeft, 
       " and on right at base position ", FwdTrimRight))
```


##Quality filter and trim
```{r quality-filter-and-trim}
#Create vector of modified names of fastq files after filtering in which the modified file name
#changes from "samplename.fastq" to "samplename.filt.fastq"
# filtFs <- str_replace(fnFs, "(.fastq)", ".filt\\1")
filtFs <- file.path(filtpathF, str_replace(basename(fnFs), "(.fastq)", ".filt\\1"))
head(filtFs)

#Filter and trim sequences listed in fnFs and store them in filtFs, while storing summary in dataframe out
out <- filterAndTrim(fwd = fnFs, filt = filtFs, rev = NULL, filt.rev = NULL,
              trimLeft = FwdTrimLeft, truncLen = c(FwdTrimRight), maxEE = 2, truncQ = 11, maxN = 0, 
              rm.phix = TRUE, compress = TRUE, verbose = TRUE, multithread = TRUE)
head(out)
#truncate file extension (".fastq" or ".fastq.gz") from rownames
rownames(out) <- rownames(out) %>% str_replace(".fast.+", "")
head(out)
```

```{r calculate-filter-trim-processing-time}
#Calculate cumulative processing time
print("Cumulative processing time (seconds):")
proc.time() - start_time
```


##Sequence inference

```{r sample-inference-1}
## Set paramaters
#Create character of all files in directory filtpathF.1 (i.e. all filtered files)
#with names including string "fastq" (i.e. all *.fastq and *.fastq.gz files)
filtFs.pass <- list.files(filtpathF, pattern="fastq", full.names = TRUE)
head(filtFs.pass)

#Create character vector of sample names from the file names in filtFs.1
#This code assumes all filename = "samplename.filt.fastq"" or "samplename.filt.fastq.gz""
sample.names.filt.pass <- sapply(strsplit(basename(filtFs.pass), ".filt.fast"), `[`, 1) 
head(sample.names.filt.pass)

#Name the elements in filtFs according to the elements in sample.names.filt
names(filtFs.pass) <- sample.names.filt.pass

#Samples for which no sequences passed the filtering step
print("Samples for which no sequences passed filter criteria (i.e. samples excluded from filtered dataset):")
sample.names.filt.fail <- setdiff(sample.names, sample.names.filt.pass)
sample.names.filt.fail

```


```{r learn-errors}
# Set random seed for reproducibility purposes
set.seed(100)

#Learn forward read error rates
#errF <- learnErrors(filtFs.pass, nread=1e6, multithread=TRUE, randomize = "TRUE")
errF <- learnErrors(filtFs.pass, nbases = 1e+09, multithread=TRUE, randomize = "TRUE")



#Plot convergence of error rate computation
plot(dada2:::checkConvergence(errF), type = "o", col = "firebrick3", main = "Convergence")

#Plot calculated errors
p.err.F <- plotErrors(errF, nominalQ = TRUE)
p.err.F

#Save figure to file
ggsave(filename = file.path(pathOut, paste0("Error_rates_fwd_", run.date, ".pdf")),
       plot = p.err.F, device = "pdf", width = 8, height = 6, units = "in")
```

```{r calculate-learn-error-processing-time}
#Calculate cumulative processing time
print("Cumulative processing time (seconds):")
proc.time() - start_time
```

```{r dereplication-and-sample-inference}
# Dereplicate and apply error rate to resolve sequence variants


#Dereplicate the filtered forward sequences by sample
derepFs <- derepFastq(filtFs.pass, verbose = FALSE)

# Set random seed for reproducibility purposes
set.seed(100)

#Run dada on the dereplicated forward sequences to resolve sequence variants
#based on the forward read error model calculated in errF
dadaFs <- dada(derepFs, err = errF, multithread = TRUE)


```
  
```{r calculate-sequence-inference-processing-time}
#Calculate cumulative processing time
print("Cumulative processing time (seconds):")
proc.time() - start_time
```
  
##Construct sequence table and remove chimeras

```{r seqtab-chimera-removal-1}
#Make sequence table and review dimensions
seqtab.prechimeraremoval <- makeSequenceTable(dadaFs)
print("Sequence table dimensions (samples, resolved sequences) before chimera removal:")
dim(seqtab.prechimeraremoval)
# Inspect distribution of sequence lengths
print("Distribution of sequence lengths before chimera removal:")
table(nchar(getSequences(seqtab.prechimeraremoval)))

# Remove chimeras
seqtab.nochim <- removeBimeraDenovo(seqtab.prechimeraremoval,
                                    method = "consensus",
                                    multithread = TRUE,
                                    verbose = TRUE)

print("Sequence table dimensions (samples, resolved sequences) after chimera removal:")
dim(seqtab.nochim)
# Inspect distribution of sequence lengths
print("Distribution of sequence lengths after chimera removal:")
table(nchar(getSequences(seqtab.nochim)))
print("Proportion of reads eliminated by chimera removal:")
sum(seqtab.nochim) / sum(seqtab.prechimeraremoval)
```

```{r save-RDS-and-image}
#Save seqtab.nochim RDS file
print("Save seqtab.nochim as RDS file named:")
f.seqtab.nochim.RDS <- file.path(pathF.RDS, paste0("MiSeq_", run.date, "_preprocessing_single_nochim.RDS"))
f.seqtab.nochim.RDS
saveRDS(seqtab.nochim, f.seqtab.nochim.RDS)

```


```{r calculate-chimera-removal-processing-time}
#Calculate cumulative processing time
print("Cumulative processing time (seconds):")
proc.time() - start_time
```

##Track reads through pipeline
 

```{r track-reads}
#Function to calculate reads for each element in dada output
getN <- function(x) sum(getUniques(x))

#Read count remaining at each step of processing. Would need additional columns if analyzing
#paired-end data.

#Create interim dataframes listing the sequence count after each step of processing per sample
print("Read count remaining at each step of processing:")
outdf <- out %>% 
  as.data.frame() %>%
  tibble::rownames_to_column() %>% 
  dplyr::rename(Sample = "rowname", input = "reads.in", filtered = "reads.out")

dadacountdf <- sapply(dadaFs, getN) %>% 
  as.data.frame() %>% 
  tibble::rownames_to_column() %>%
  dplyr::rename(Sample = "rowname", denoisedF = ".")

nonchimdf <- rowSums(seqtab.nochim) %>% 
  as.data.frame() %>% 
  tibble::rownames_to_column() %>%
  dplyr::rename(Sample = "rowname", nonchim = ".")

#Join dataframes by Sample column to track counts for each sample at each step of processing
track <- dplyr::left_join(outdf, dadacountdf, by = c("Sample" = "Sample")) %>%
  dplyr::left_join(nonchimdf, by = c("Sample" = "Sample"))
head(track)

#Percentage of original unfiltered reads remaining at each step of processing
print("Percentage of original unfiltered foward reads remaining at each step of processing:")
track.percent <- dplyr::mutate_at(track, vars(-matches("Sample")), funs(. * 100 / track[["input"]]))
track.percent <- rename_at(track.percent, vars(-matches("Sample")), funs(paste0(., "_percent")))
head(track.percent)

#Create a "long" form version of track.percent for use with plotting
track.percent.long <- track.percent %>% 
  tidyr::gather(key = "Step", value = "Percent", -Sample) 

#Modify track.percent.long so that Step column is a factor with levels in the correct order
track.percent.long[["Step"]] <- track.percent.long[["Step"]] %>% forcats::fct_relevel(colnames(track.percent[-1]))

#Plot percentages of input forward reads remaining after each step on a per sample basis.
p.track <- ggplot(track.percent.long, aes(x = Step, y = Percent)) + 
  geom_line(aes(group = Sample, colour = Sample)) +
  geom_point() +
  labs(title = "Percent of input forward reads remaining\nafter each processing step per sample",
       subtitle = paste0("Fwd read left trim position: ", FwdTrimLeft, "\nFwd read right trim position: ", FwdTrimRight),
       x = "Processing Step", y = "% of Input Reads Remaining") +
  theme(plot.title = element_text(hjust = 0.5, size = 18), 
        plot.subtitle = element_text(hjust = 0.5, size = 16),
        legend.position = "none",
        axis.text = element_text(size = 12), axis.title = element_text(size = 16))
p.track

#Save figure to file
ggsave(filename = file.path(pathOut, paste0("read_track_fwd_L", FwdTrimLeft, "_R", FwdTrimRight, "_", run.date, ".pdf")), 
       plot = p.track, device = "pdf", width = 8, height = 6, units = "in")

#Merge track and track.percent dataframes
track.merged <- dplyr::left_join(track, track.percent, by = c("Sample" = "Sample"))
head(track.merged)

#Save log file of tracking reads to file.
print("Save track.merged dataframe tracking read counts as TSV file named:")
f.track.merged <- file.path(pathOut, paste0("MiSeq_", run.date, "_fwd_read_tracking_log.txt"))
head(f.track.merged)
readr::write_tsv(track.merged, f.track.merged)

```


##Assign taxonomy

```{r assign-taxonomy}


# RDP
#Record start time for this taxonomy assignment step
tax_start_time = proc.time()
print(paste0("RDP training database: ", "rdp_train_set_16.fa.gz"))
taxa.rdp <- assignTaxonomy(seqtab.nochim, file.path(pathdb, "rdp_train_set_16.fa.gz"), multithread = TRUE) 

#Add RDP species assignment
print(paste0("RDP species assignment database: ", "rdp_species_assignment_16.fa.gz"))
taxa.rdp.plus <- addSpecies(taxa.rdp, file.path(pathdb, "rdp_species_assignment_16.fa.gz"))
writeLines("RDP taxonomy assignment step processing time (seconds):")
proc.time() - tax_start_time


```


```{r calculate-taxonomy-assignment-processing-time}
#Calculate cumulative processing time
print("Cumulative processing time (seconds):")
proc.time() - start_time
```

##Generate PhyloSeq Objects and final save

```{r create-phyloseq}
# Create PhyloSeq objects

# Load mapping file for sequencing run
print("Sequencing run mapping file:")
f.mapping <- file.path(pathMap, list.files(path = pathMap, pattern = ".txt$"))
f.mapping
mapping <- readr::read_tsv(file = f.mapping, col_names = TRUE)

#Remove any rows with "#SampleID" column value NA and convert to dataframe
mapping <- mapping %>%
  dplyr::filter(!is.na(`#SampleID`)) %>%
  as.data.frame()

#Convert mapping$#SampleID column to rownames
mapping <- tibble::column_to_rownames(mapping, var = "#SampleID")


# RDP
#ps0.rdp <- phyloseq(otu_table(seqtab.1, taxa_are_rows = FALSE), tax_table(taxa.rdp.plus), phy_tree(tree))
ps0.rdp <- phyloseq(otu_table(seqtab.nochim, taxa_are_rows = FALSE), 
                    tax_table(taxa.rdp.plus),
                    sample_data(mapping))
ps0.rdp



# Sanity checks
get_taxa_unique(ps0.rdp, "Phylum")


# Save RDS files for downstream analysis

saveRDS(ps0.rdp, file = file.path(pathps0, paste0("ps0.", run.date, ".rdp_v1.RDS")))


#Save current R image (workspace) as .RData file
print("Save R image (workspace) as .RData file named:")
f.image <- file.path(pathF.RDS, paste0("MiSeq_", run.date, "_preprocessing_single.RData"))
f.image
save.image(f.image)
```

```{r calculate-final-processing-time}
#Calculate total processing time
print("Total processing time (seconds):")
proc.time() - start_time
```
